Formal Synthesis of Digital Controllers for Observable Continuous Plants with Transient Performance Specifications

### 2018-01-24 TACAS2018/paper.tex
ch 4

%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Modelling Errors in the Control Loop}\label{sec:errors}
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

@@

A common occurrence in digital systems is that they do not necessarily
satisfy the real-time specifications.  Delays due to processing time,
communication lines, or other mechanisms will cause disturbances in the
control loop.  In our model these delays are all deferred to the ZOH, i.e.,
we presume that everything that happens in the digital world takes place at
the sampling instance $t$ and the output of the result is delayed at the DAC
when it is fed back into the real world.  We~will also presume that this
delay is much smaller than the sampling time.

\begin{theorem}
The error introduced by a delay $t_d$ in the control output $\vec{u}_k$ of a discrete feedback controller for a continuous LTI system can be modelled as a non-deterministic noise signal upper bound by 
\begin{align}
\norm{\vec{y}_{k}-{\vec{y}_{t_d}}_{k}}_1 \leq q_t, \text{ where } q_t=&(\norm{\mat{C}(\mat{B}_{T_s}-\mat{B}_{t_s})}_1+\norm{\mat{C}\mat{A}_{t_s}\mat{B}_{t_d}}_1)\norm{U}_1
\end{align}
where $\vec{y}_{t_d}$ is the output of the system with delay $t_d$, and $\norm{\vec{x}}_1=\sum_{i=0}^{size(\vec{x})} |\vec{x}_i|$ is the $l_1$ norm (in for sets, $\norm{X}_1 = sup \{\norm{\vec{x}}_1 \mid \vec{x} \in X\}$). 
\end{theorem}

@@

In Figure \ref{fig:observersystem2} we introduced several sources of noise,
which are due to quantization effects.  These effects are caused by the use
of imprecise numerical representations over the reals in the digital domain
of the controller which is an important consideration when trying to synthesise
sound controllers.
The models we consider have two sources of error that are due to numerical  
representation which we commonly refer to as FWL effects.
The first is the numerical error introduced by the
fixed-point numbers employed to model the plant, i.e., to represent the
plant dynamics $A_d$, $B_d$, $C_d$, $D_d$, $x_k$ and $y_k$.
The second is the quantization error
introduced by the digital controller, which performs operations on
fixed-point numbers.  In this section we outline the notation for the
representation of numbers, and briefly describe the errors
introduced.  A~formal discussion is in
Appendix~\ref{sec:numeric_rep}.

Let $\mathbb{R}_{\langle M,F \rangle}$ be a fixed point domain with $M$ bits
representing the binary word length of the representation and $F$ bits
representing the prescaling $2^{-F}$ (\emph{i.e} the number of bits after
the decimal place).  We note that a negative $F$ denotes an integer
multiplication that adds $-F$ trailing zeros in the number represented by
$M$.

We will use the domain $\mathbb{R}_{\langle C \rangle} \text{ where } \langle C \rangle = \langle M_c,F_c \rangle$
to represent the fixed point precision of the controller, and 
$\mathbb{R}_{\langle P \rangle} \supseteq \mathbb{R}_{\langle C \rangle}$
to represent the domain of a digital model of the dynamics of a real plant when
evaluated by a verification tool.
Thus, any mathematical operations in our modelled digital controller will be executed in the
domain $\mathbb{R}_{\langle C \rangle}$, and all other calculations
in our model will be carried out in the domain $\mathbb{R}_{\langle P \rangle}$.
Given ${\langle M,F \rangle}$ we use $\mathcal{F}_{\langle M,F \rangle} : \mathbb{R} \rightarrow \mathbb{R}_{\langle M,F \rangle}$ as the truncation function.

@@

Since we are interested in synthesising controllers with observer
designs, in this section we will look at checking stability, safety
and the response time in the presence of an observer. \footnote{A full description of
the theory related to the observer design is provided in appendix \ref{sec:control_theory}.}

@@

Discretization of a continuous dynamical system turns the ODE into a difference
equation, assuming zero-order hold for the input $\vec{u}$ (i.e. the value $u(t)$
 is fixed for the integration period), %and continuous integration for the noise $\vec{w}$, 
into
\begin{align}
\label{eq:discretization}
\vec{x}_{k+1} &= \mat{A}_d\vec{x}_k+\mat{B}_d\vec{u}_k\\% + \vec{w}_k\\
y_k &= \mat{C}_d \vec{x}_ k + \mat{D}_d \vec{u}_ k 
\end{align}
%with covariance for $\vec{w}_k$, $N ( 0 , Q_d )$
where
\begin{align}
\label{eq:discretize}
\mat{A}_d &= e^{\mat{A} T_s} \\%= \mathcal{L}^{-1} { ( s \mat{I} - \mat{A} )^{-1} }_{t = T_s}\\
\mat{B}_d &= \int_{t = 0}^{T_s} e^{\mat{A} t} dt\ \mat{B} = \mat{A}^{-1} ( \mat{A}_d - \mat{I} ) \mat{B}\\
\mat{C}_d &= \mat{C}\\
\mat{D}_d &= \mat{D} %\\
%Q_d &= \int_{t = 0}^{T_s} e^{\mat{A} t} Q e^{\mat{A}^T t} dt
\end{align}
and $T_s$ is the sample time. Then
\begin{align*}
\forall k,  \vec{x}(kT_s)=\vec{x}_k \text{ and } \vec{y}(kT_s) = \vec{y}_k
\end{align*}
%Let $g_d(k)=\mathcal{G}(t,g(t),T)$ be a function that performs the discretization described above, where $g(t)$
%represents the continuous dynamics and $g_d(k)$ the corresponding discrete dynamics. 
%Let $G(s)=\mathcal{L}(g(t))$ and $G_d(z)=\mathcal{Z}(g_d(k))$ be the corresponding Laplace and Z-transforms
%of $g(t)$ and $g_d(k)$. Given this relation, we have 
%$$G_d(z)=G(z)|_{z=e^{sT}} \text{ where } g_d(k)=\mathcal{G}(t,g(t),T) \wedge T < \frac{1}{.5f_s}$$
%Where $.5f_s$ is the Nyquist frequency of $g(t)$. This last restriction is introduced to avoid the effects of aliasing
%which could cause non-existent `phantom' poles to appear otherwise at fractions of the true frequency they represent. 
The eigenvalues of $\vec{A}_d$ %corresponding to the poles of $G_d(z)$,
and those of $\vec{A}$ %corresponding to the poles of $G(s)$
are related as follows:
$\hat{\lambda}_i=e^{\lambda_iT_s} \text{ where } \hat{\lambda}_i \in \sigma(\vec{A}_d) \text{ and } \lambda_i \in \sigma(\vec{A})$
where $\sigma(\cdot)$ is the spectrum of a matrix.
The following remark is worth mentioning regarding the above.
\begin{remark}
The witnessed maximum amplitude of the discrete signals $\vec{x}_k,\vec{y}_k$
may be smaller to that of $\vec{x}(t),\vec{y}(t)$ due to synchronism at
fraction-frequency sampling. This means that safety verification of the state space
must account for this condition.
\end{remark}
Imagine that a signal $\vec{x}(t)=sin\left({\omega_f t}\right)$. If we sample
this signal at intervals $T_s=\frac{\pi}{\omega_f}$ starting at $t=0$, then all
measured samples will be $0$.  For this reason, systems must be sampled at
intervals smaller than the maximum composing frequency of the source signal
$T_s<\frac{\pi}{\omega_max}$.  This is the so called Nyquist Criterion~\cite{Astrom08}.  
The maximum amplitude of the original signal can be upper bounded from the
observed values using the inverse cosine of the sampling function.
Since equation \eqref{eq:discretization} models \eqref{eq:dynamical}, we may use the semantics defined in Section
\ref{sec:model_semantics} to analyse continuous dynamical systems.

@@

 Let $\mat{J}$ be a diagonal matrix, such that 
 $${\lambda_1}_i=e^{\lambda_i T_1}, \forall \lambda_i \in \mat{J}.$$
 Let us now take a different sample rate $T_2$, such that 
 $${\lambda_2}_i=e^{\lambda_i T_2}, \forall \lambda_i \in \mat{J}.$$
 We can then say that 
 \begin{equation}
 {\lambda_2}_i=e^{\lambda_i T_1 \frac{T_2}{T_1}}={\lambda_1}_i^{\frac{T_2}{T_1}} \Rightarrow A_2=A_1^{\frac{T_2}{T_1}}.
 \end{equation}
 This proves the theorem for $gm(\lambda_i)=1$.
 We now note that given an exponentiated Jordan form with geometric multiplicity, the upper diagonal terms can be
 referenced to the original eigenvalues modified by the sampling rate.
  Let $\mat{J}\in \mathbb{R}^{p \times p}$ be a Jordan Canonical matrix whose powers are described in Equation \eqref{jord:pow}
 Then
\begin{align}
 \mat{J}_1&=\sum_{k=0}^\infty \frac{1}{k!}\left(\mat{J}T_1\right)^k\nonumber\\
 &=\sum_{k=0}^\infty \frac{1}{k!} \left [ \begin{array}{cccc}
 \lambda_i^k  & \binom{k}{1}  \lambda^{k-1} & \hdots  & \binom{k}{p-1} \lambda_i^{k-p+1} \\
& \lambda_i^k  & \binom{k}{1}  \lambda_i^{k-1} & \vdots \\
\vdots & \ddots & \ddots & \vdots \\
& &  &\lambda_i^k \\
\end{array} \right ] T_1^k
\end{align}
The eigenvalues remain the same as in the diagonal case, but the upper triangular terms are of the form:
\begin{align}
\forall j>i, c_{ij}&=\sum_{k=j-i}^\infty \frac{1}{k!}\binom{k}{j-i} \lambda_i^{k-(j-i)}T_1^k\\
&=\frac{1}{(j-i)!}\sum_{k=j-i}^\infty \frac{1}{(k-(j-i))!} \lambda_i^{k-(j-i)}T_1^k\nonumber\\
&=\frac{T_1^{j-i}e^{\lambda_i T_1}}{(j-i)!}\nonumber
\end{align}
which completes the proof for $gm(\lambda_i)>1$.
\end{proof}

>>>>

%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Modelling errors in the control loop}\label{sec:errors}
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

A common occurrence in digital systems is that they do not necessarily satisfy real-time specifications, 
in view of many errors introduced in the dynamics. 
Errors might results from delays due to processing time,
corruptions over communication lines in the control loop, 
approximations due to digitalisation.

@@

In our model delays are all deferred to the ZOH \note{acronym undefined - define these and others, with reference to the Figure 1 above}, 
i.e., we presume that \note{qualitative: everything that happens in the digital world takes place at
the sampling instance $t$} and the output of the result is delayed at the DAC 
when it is fed back into the real world.  
We will also presume that this delay is much smaller than the sampling time \note{does this assumption make practical sense? why? add reference?}.

\begin{theorem}
The error introduced by a delay $t_d$ in the control output $\vec{u}_k$ of a discrete feedback controller for a continuous LTI system can be modelled as a non-deterministic noise signal upper bound by 
\begin{align}
\norm{\vec{y}_{k}-{\vec{y}_{t_d}}_{k}}_1 \leq q_t, \text{ where } q_t=&(\norm{\mat{C}(\mat{B}_{T_s}-\mat{B}_{t_s})}_1+\norm{\mat{C}\mat{A}_{t_s}\mat{B}_{t_d}}_1)\norm{U}_1
\end{align}
where $\vec{y}_{t_d}$ is the output of the system with delay $t_d$, and $\norm{\vec{x}}_1=\sum_{i=0}^{size(\vec{x})} |\vec{x}_i|$ is the $l_1$ norm (\note{what? unclear?} in for sets, $\norm{X}_1 = sup \{\norm{\vec{x}}_1 \mid \vec{x} \in X\}$). 
\end{theorem}

@@

In Figure \ref{fig:observersystem2} we introduced several sources of noise,
which are due to quantization effects.  These effects are caused by the use
of imprecise numerical representations over the reals in the digital domain
of the controller which is an important consideration when trying to synthesise
sound controllers.
The models we consider have two sources of error that are due to numerical  
representation which we commonly refer to as FWL \note{undefined acronym} effects.
The first is the numerical error introduced by the
fixed-point numbers employed to model the plant, i.e., to represent the
plant dynamics $A_d$, $B_d$, $C_d$, $D_d$, $x_k$ and $y_k$.
The second is the quantization error
introduced by the digital controller, which performs operations on
fixed-point numbers.  In this section we outline the notation for the
representation of numbers, and briefly describe the errors
introduced.  A~formal discussion is in
Appendix~\ref{sec:numeric_rep}.

Let $\mathbb{R}_{\langle M,F \rangle}$ be a fixed point domain with $M$ bits
representing the binary word length of the representation and $F$ bits
representing the prescaling $2^{-F}$ (\emph{i.e} the number of bits after
the decimal place).  We note that a negative $F$ denotes an integer
multiplication that adds $-F$ trailing zeros in the number represented by
$M$. 
We will use the domain $\mathbb{R}_{\langle C \rangle} \text{ where } \langle C \rangle = \langle M_c,F_c \rangle$
to represent the fixed point precision of the controller, and 
$\mathbb{R}_{\langle P \rangle} \supseteq \mathbb{R}_{\langle C \rangle}$
to represent the domain of a digital model of the dynamics of a real plant when
evaluated by a verification tool.
Thus, any mathematical operations in our modelled digital controller will be executed in the
domain $\mathbb{R}_{\langle C \rangle}$, and all other calculations
in our model will be carried out in the domain $\mathbb{R}_{\langle P \rangle}$.
Given ${\langle M,F \rangle}$ we use $\mathcal{F}_{\langle M,F \rangle} : \mathbb{R} \rightarrow \mathbb{R}_{\langle M,F \rangle}$ as the truncation function.

@@

Since we are interested in synthesising controllers with observers, 
in this section we will look at checking stability, safety and the response time in the presence of an observer. \footnote{A full description of
the theory related to the observer design is provided in appendix \ref{sec:control_theory}.}

@@

Discretization of a continuous dynamical system turns the ODE into a difference
equation, assuming zero-order hold for the input $\vec{u}$ (i.e. the value $u(t)$
 is fixed for the integration period), %and continuous integration for the noise $\vec{w}$, 
into
\begin{align}
\label{eq:discretization}
\vec{x}_{k+1} &= \mat{A}_d\vec{x}_k+\mat{B}_d\vec{u}_k\\% + \vec{w}_k\\
y_k &= \mat{C}_d \vec{x}_ k + \mat{D}_d \vec{u}_ k 
\end{align}
%with covariance for $\vec{w}_k$, $N ( 0 , Q_d )$
where
\begin{align}
\label{eq:discretize}
\mat{A}_d &= e^{\mat{A} T_s} \\%= \mathcal{L}^{-1} { ( s \mat{I} - \mat{A} )^{-1} }_{t = T_s}\\
\mat{B}_d &= \int_{t = 0}^{T_s} e^{\mat{A} t} dt\ \mat{B} = \mat{A}^{-1} ( \mat{A}_d - \mat{I} ) \mat{B}\\
\mat{C}_d &= \mat{C}\\
\mat{D}_d &= \mat{D} %\\
%Q_d &= \int_{t = 0}^{T_s} e^{\mat{A} t} Q e^{\mat{A}^T t} dt
\end{align}
and $T_s$ is the sample time. Then
\begin{align*}
\forall k,  \vec{x}(kT_s)=\vec{x}_k \text{ and } \vec{y}(kT_s) = \vec{y}_k
\end{align*}
%Let $g_d(k)=\mathcal{G}(t,g(t),T)$ be a function that performs the discretization described above, where $g(t)$
%represents the continuous dynamics and $g_d(k)$ the corresponding discrete dynamics. 
%Let $G(s)=\mathcal{L}(g(t))$ and $G_d(z)=\mathcal{Z}(g_d(k))$ be the corresponding Laplace and Z-transforms
%of $g(t)$ and $g_d(k)$. Given this relation, we have 
%$$G_d(z)=G(z)|_{z=e^{sT}} \text{ where } g_d(k)=\mathcal{G}(t,g(t),T) \wedge T < \frac{1}{.5f_s}$$
%Where $.5f_s$ is the Nyquist frequency of $g(t)$. This last restriction is introduced to avoid the effects of aliasing
%which could cause non-existent `phantom' poles to appear otherwise at fractions of the true frequency they represent. 
The eigenvalues of $\vec{A}_d$ %corresponding to the poles of $G_d(z)$,
and those of $\vec{A}$ %corresponding to the poles of $G(s)$
are related as follows:
$\hat{\lambda}_i=e^{\lambda_iT_s} \text{ where } \hat{\lambda}_i \in \sigma(\vec{A}_d) \text{ and } \lambda_i \in \sigma(\vec{A})$
where $\sigma(\cdot)$ is the spectrum of a matrix.
The following remark is worth mentioning regarding the above.
\begin{remark}
The witnessed maximum amplitude of the discrete signals $\vec{x}_k,\vec{y}_k$
may be smaller to that of $\vec{x}(t),\vec{y}(t)$ due to synchronism at
fraction-frequency sampling. This means that safety verification of the state space
must account for this condition.
\end{remark}
Imagine that a signal $\vec{x}(t)=sin\left({\omega_f t}\right)$. If we sample
this signal at intervals $T_s=\frac{\pi}{\omega_f}$ starting at $t=0$, then all
measured samples will be $0$.  For this reason, systems must be sampled at
intervals smaller than the maximum composing frequency of the source signal
$T_s<\frac{\pi}{\omega_max}$.  This is the so called Nyquist Criterion~\cite{Astrom08}.  
The maximum amplitude of the original signal can be upper bounded from the
observed values using the inverse cosine of the sampling function.
Since equation \eqref{eq:discretization} models \eqref{eq:dynamical} \note{fix ref}, we may use the semantics defined in Section
\ref{sec:model_semantics} to analyse continuous dynamical systems.

@@

 Let $\mat{J}$ be a diagonal matrix, such that 
 $${\lambda_1}_i=e^{\lambda_i T_1}, \forall \lambda_i \in \mat{J}.$$
 Let us now take a different sample rate $T_2$, such that 
 $${\lambda_2}_i=e^{\lambda_i T_2}, \forall \lambda_i \in \mat{J}.$$
 We can then say that 
 \begin{equation}
 {\lambda_2}_i=e^{\lambda_i T_1 \frac{T_2}{T_1}}={\lambda_1}_i^{\frac{T_2}{T_1}} \Rightarrow A_2=A_1^{\frac{T_2}{T_1}}.
 \end{equation}
 This proves the theorem for $gm(\lambda_i)=1$.
 We now note that given an exponentiated Jordan form with geometric multiplicity, the upper diagonal terms can be
 referenced to the original eigenvalues modified by the sampling rate.
  Let $\mat{J}\in \mathbb{R}^{p \times p}$ be a Jordan Canonical matrix whose powers are described in Equation \eqref{jord:pow}
 Then
\begin{align}
 \mat{J}_1&=\sum_{k=0}^\infty \frac{1}{k!}\left(\mat{J}T_1\right)^k\nonumber\\
 &=\sum_{k=0}^\infty \frac{1}{k!} \left [ \begin{array}{cccc}
 \lambda_i^k  & \binom{k}{1}  \lambda^{k-1} & \hdots  & \binom{k}{p-1} \lambda_i^{k-p+1} \\
& \lambda_i^k  & \binom{k}{1}  \lambda_i^{k-1} & \vdots \\
\vdots & \ddots & \ddots & \vdots \\
& &  &\lambda_i^k \\
\end{array} \right ] T_1^k
\end{align}
The eigenvalues remain the same as in the diagonal case, but the upper triangular terms are of the form:
\begin{align}
\forall j>i, c_{ij}&=\sum_{k=j-i}^\infty \frac{1}{k!}\binom{k}{j-i} \lambda_i^{k-(j-i)}T_1^k\\
&=\frac{1}{(j-i)!}\sum_{k=j-i}^\infty \frac{1}{(k-(j-i))!} \lambda_i^{k-(j-i)}T_1^k\nonumber\\
&=\frac{T_1^{j-i}e^{\lambda_i T_1}}{(j-i)!}\nonumber
\end{align}
which completes the proof for $gm(\lambda_i)>1$. \qed
\end{proof}
